---
title: "Estimating the cut-off for the overlap of two ranked gene lists"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating the cut-off for the overlap of two ranked gene lists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 7.5)
library(thresholdopt)
```

## The problem

Two studies of the same perturbation publish two lists of differentially
expressed genes, and the size of their overlap depends almost entirely on
where each study cut its list — an adjusted p-value, a fold-change floor and
a fixed top-*k* are not comparable across experiments. `thresholdopt`
estimates the cut-off from the two rankings instead of assuming it.

## Usage with named vectors

The input is two named numeric vectors: names are the gene identifiers,
values are the ranking statistic (a log2 fold change, a *t* statistic, any
continuous score — only `abs(value)` is used for ranking). The bundled data
are the two SF-1 overexpression experiments of the source publication.

```{r vectors}
data(sf1_overexpression)
lalli  <- setNames(sf1_overexpression$doghman$logFC,
                   sf1_overexpression$doghman$Symbol)
ferraz <- setNames(sf1_overexpression$ferraz$logFC,
                   sf1_overexpression$ferraz$Symbol)

fit <- threshold_overlap(lalli, ferraz,
                         label.x = "Lalli's SF1 Overexpression",
                         label.y = "Ferraz's SF1 Overexpression")
```

The threshold (3221) and the intersection (1096) are the values reported in
the source publication, and the figure is the publication's diagnostic plot:
the two ranked lists with the selected prefix in green, and the objective
function with the intersection size on the secondary axis, the optimum
marked in green.

The returned object carries everything shown, plus the consensus:

```{r object}
str(fit, max.level = 1)
head(fit$consensus)
```

## Usage with differential-expression tables

`threshold_overlap_deg()` accepts the data frames produced by
differential-expression pipelines; the defaults match `DESeq2::results()`
(identifiers in the row names, statistic in `log2FoldChange`, `NA` rows
dropped). When labels are omitted, the names of the objects are used.

```{r deg}
lalli_df  <- data.frame(log2FoldChange = sf1_overexpression$doghman$logFC,
                        row.names = sf1_overexpression$doghman$Symbol)
ferraz_df <- data.frame(log2FoldChange = sf1_overexpression$ferraz$logFC,
                        row.names = sf1_overexpression$ferraz$Symbol)
fit2 <- threshold_overlap_deg(lalli_df, ferraz_df)
identical(fit2$threshold, fit$threshold)
```

## How the optimum is found

For a common universe of $N$ features and every candidate size $n$: the
intersection $q(n)$ of the two top-$n$ sets is scored with a hypergeometric
tail probability $p(n)$, evaluated in log space so genome-scale lists do not
saturate at the double-precision underflow limit near $10^{-322}$, and
Bonferroni-adjusted over the sweep. The score alone cannot select a size — a
list against itself gives $p = 0$ everywhere, a list against its own reverse
is minimised only at the extremes — so it is regularised by the divergence
$d(n) = n - q(n)$:

$$\lambda(n) = \frac{d(n)}{\max_n d(n)}, \qquad
  O(n) = \lambda(n)\,\log_{10} p_{adj}(n),$$

and the selected threshold minimises $O(n)$. The divergence vanishes exactly
where the raw score degenerates, which is what makes the optimum interior
and well defined. The whole sweep costs $O(N \log N)$.

## What the numbers do and do not claim

The optimiser always returns a threshold, including for unrelated lists: the
objective is a selection score, not a calibrated test of the selected
result. Interpreting the overlap — is it significant, and is the shared
response directionally coherent? — is a separate inferential problem,
addressed in the development version of this package.

## Reference

Muzzi JCD, Ruggiero C, Doghman-Bouguerra M, et al. Steroidogenic factor-1
regulates a core set of target genes to promote malignancy in adrenocortical
carcinoma. *European Journal of Endocrinology* 193(1):135-145, 2025.
doi:10.1093/ejendo/lvaf138
